Comparing software variability across and within fMRI analysis packages

### 2021-12-23 paper/paper.tex

%%%%%%%%%% Var. of Thresh %%%%%%%%
\begin{figure*}[ht]
  \begin{subfigure}[ht]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/act_deact-marked.pdf}
    % \includegraphics[width=\textwidth]{figures/plots/wt-tstats/act_deact-framed.png}
    % \includegraphics[width=\textwidth]{figures/abs/gl-thresh.png}
    %\caption{Standard deviation of thresholded t-statistics map on template surface}
  \end{subfigure}
  \centering
  \caption{Thresholded group-level t-statistics within tools at machine error for FSL (\textbf{A}),
    SPM (\textbf{B}), AFNI (\textbf{C}), and confusion matrices of activation instability
    in BT and WT among the 360 regions of the HCP-MMP1.0 parcellation (\textbf{D}).}
  \label{fig:thresh-maps}
\end{figure*}

@@

In fMRI group analyses, within-tool variability remains an order of magnitude smaller
than between-tool variability. Group analyses benefit from regularization
of numerical noise, which is expected to increase as sample size
increases. Therefore, for fMRI studies with large sample sizes, within-tool variability may be negligible with respect to between-tool variability. The
recommendation made in~\cite{botvinik2020variability} to rely on
``multiverse'' analyses where multiple analysis tools are compared is
therefore likely to successfully correct for machine error in group
studies. Nevertheless, within-tool variability remains substantial in group analyses
that are based on a single tool, as is commonly the case in current fMRI
studies. In particular, in our study, the inherent instability of
thresholding was triggered by machine error in 9\% of 360 brain regions,
which indicates that machine error might have impacted neuroscientific
conclusions related to these regions.

In subject-level analyses, within-tool and between-tool variabilty can become
of comparable magnitude for some subjects in some regions. This observation
is particularly relevant to the development of fMRI-based biomarkers aiming
at individualized phenotype predictions. Machine error is expected to play
a non-negligible role in such analyses, even when predictions combine
results produced by multiple tools. The observed regularization effect of
group analyses toward numerical noise is consistent with observations made
in~\cite{kiar2020numerical} from diffusion MRI where connectome graph
statistics were found to be substantially unstable at the subject-level
while group distributions remained consistent. \gk{this last sentence feels like it belongs in the previous paragraph, right before mentioning thresholding}

For both group- and subject-level analyses, between-tool variability and within-tool
variability were found to be moderately correlated. Even though between-tool
variability and within-tool  variability are different in nature, this result
suggests that in some cases both sources of variability may have a common
cause that might be related to the conditioning of the BOLD signal
estimation problem \gk{I wouldn't necessarily say it's because of "BOLD signal estimation problem", just "algorithms used to process"} in a specific dataset. Therefore, in some regions,
instabilities of similar magnitude may be triggered by small numerical
perturbations, model variations, or implementation differences. For
instance, the analysis of datasets with high motion may be unstable both
across tools and numerically.

Therefore, numerical stability may be a suitable proxy to study
between-tool variability. This speculation might be of practical value to
address software variability at large given that numerical stability refers
to a consistent mathematical framework whereas between-tool variability
remains more empirical. Likewise, useful quality control metrics may be derived from numerical and between-tool stability.

The finding that numerical variability approached tool variability at the
virtual precision of t=17~bits is interesting too. Indeed, while machine
error generally introduces differences in the order of \yohanmod{1} \yohan{1/2} ulp
\yohan{Are we talking about one arithmetic operation? Or an entire execution?}
--- or
t=53~bits for double-precision values and t=24~bits for single-precision
ones \yohan{I would not mix virtual precision and error measurement}--- common scientific software dependencies introduce larger errors
.
For instance, SciPy's 2D spline interpolation was recently found to be
precise up to t=10~bits~\cite{pytracer} \tristan{Fix ref}
\yohan{Same as previous, I would talk about significance or relative error, or remove the t= and not talk about virtual precision} and might therefore introduce
numerical perturbations leading to errors in the range of between-tool
variability \tristan{Yohan, please check that sentence}. Such high errors
might be triggered by updates in operating systems, Python, MATLAB, and
other software dependencies.
\yohan{The sentence is ok, just a picky point: the SciPy result you cited describes the precision of the spline implementation while at the end you're talking differences due to updates.
  The nuance is that two versions may have a similar precision and so when you udpate your
  system, the difference between both versions is low but each version can be ill-conditionned.}

Our results are limited by the type of numerical noise introduced in the
analyses. We only perturbed the outputs of elementary
mathematical functions while numerical noise could creep in any
floating-point operation. Our estimation of numerical variability should
therefore be considered a lower bound. Our estimation of tool varability is
also likely to be underestimated, having tested only 3 analytical pipelines
among the thousands available~\cite{carp2012plurality}.

In conclusion, our results motivate further numerical stability
investigations in fMRI analysis pipelines. Pipeline-level analyses could be
conducted to identify specific components that contribute to numerical
variability, and if possible correct them accordingly. When instability is
inherent to the analysis, sampling results distributions through numerical
perturbations might improve stability, as explored in~\cite{kiar2021data}.
Finally, pipeline-specific statistical corrections might be envisaged to
account for numerical variability.

>>>>

%%%%%%%%%% Var. of Thresh %%%%%%%%
\begin{figure*}[ht]
  \begin{subfigure}[ht]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/act_deact-marked.pdf}
    % \includegraphics[width=\textwidth]{figures/plots/wt-tstats/act_deact-framed.png}
    % \includegraphics[width=\textwidth]{figures/abs/gl-thresh.png}
    %\caption{Standard deviation of thresholded t-statistics map on template surface}
  \end{subfigure}
  \centering
  \caption{\textbf{A},\textbf{B},\textbf{C}: Thresholded group-level
    t-statistics within tools at machine error for FSL (\textbf{A}), SPM
    (\textbf{B}) and AFNI (\textbf{C}). Arrows point to activation clusters
    impacted by both within- and between-tool variability \tristan{Ali: is it really the case? Can you make the arrows point more accurately to 
    the clusters? Some of them are quite far from the cluster they point to.}.
     \textbf{D}: Confusion matrices of
    activation instability in BT and WT among the 360 regions of the
    HCP-MMP1.0 parcellation.}
  \label{fig:thresh-maps}
\end{figure*}

@@

In fMRI group analyses, within-tool software variability remains an order
of magnitude smaller than between-tool variability. This is likely
explained by the fact that group analyses benefit from regularization of
numerical noise, which is expected to increase with sample size. This finding is
consistent with observations made in~\cite{kiar2020numerical} from
diffusion MRI data where connectome graph statistics were found to be
substantially unstable at the subject level while group distributions
remained consistent. Therefore,
for fMRI studies with large sample sizes, within-tool software variability
may be neglected with respect to between-tool variability. In particular,
multiverse analyses aggregating the outcome of multiple analysis tools are
likely to successfully correct for machine error in such group studies.  Nevertheless, within-tool software variability remains
substantial in group analyses that are based on a single tool, as is
commonly the case in current fMRI studies. In particular, in our study, the
inherent instability of thresholding was triggered by within-tool software
variability in 9\% of 360 brain regions, which indicates that it might have
impacted neuroscientific conclusions related to these regions.

In subject-level analyses, within- and between-tool software variabilties can become
of comparable magnitude for some subjects in some regions. This observation
is particularly relevant to the development of fMRI-based biomarkers aiming
at individualized phenotype predictions. Machine error may play
a non-negligible role in such analyses, even when predictions combine
results produced by multiple tools.

For both group- and subject-level analyses, between- and within-tool
software variabilities were found to be moderately correlated. Even though
both types of variability are different in nature, this result suggests
that in some cases they may have a common cause that might be related to
the conditioning of fMRI analysis in a specific dataset. Along these lines,
in some regions, instabilities of similar magnitude may be triggered by
small numerical perturbations, model variations, or implementation
differences. For instance, high-motion datasets may trigger between- and
within-tool instabilities similarly.

Therefore, numerical stability may be a suitable proxy to study
between-tool variability. This speculation might be of practical value to
address software variability at large given that within-tool variability
measurements can be automated while between-tool ones require substantial
human intervention. Indeed, within-tool variability can be simulated
irrespective of the particular analysis done, and with important sample
sizes, while between-tool variability requires re-implementing analysis in
different toolboxes and is limited to a few samples. Automated multi-verse
frameworks could be developed to assess and possibly reduce within-tool
variability as a pre-requisite for between-tool variability.

\tristan{This paragraph needs to be revised as we can't conclude that from the data we have}
The finding that numerical variability approached tool variability at the
virtual precision of t=17~bits is interesting too. Indeed, while machine
error generally introduces perturbations on the least-significant bit of the mantissa,
ones 
% \yohan{I would not mix virtual precision and error measurement}\tristan{better?}
common scientific software dependencies introduce larger perturbations
.
For instance, SciPy's 2D spline interpolation was recently found to be
precise up to 10~bits~\cite{pytracer}
and its replacement by a more precise implementation might therefore introduce
numerical perturbations leading to errors in the range of between-tool
variability. Such high errors
might be triggered by updates in operating systems, Python, MATLAB, and
other software dependencies.
% \yohan{The sentence is ok, just a picky point: the SciPy result you cited
% describes the precision of the spline implementation while at the end
% you're talking differences due to updates. The nuance is that two versions
% may have a similar precision and so when you udpate your system, the
% difference between both versions is low but each version can be
% ill-conditionned.}
% \tristan{better?}

Our results are limited by the type of numerical noise introduced in the
analyses. Indeed, we only perturbed the outputs of elementary
mathematical functions while numerical noise could creep in any
floating-point operation. Therefore, our estimation of within-tool software variability should
be considered a lower bound. Likewise, our estimation of tool varability is
likely to be underestimated, having tested only 3 analytical pipelines
among the thousands available~\cite{carp2012plurality}.

In conclusion, our results motivate further numerical stability
investigations in fMRI analyses. Pipeline-level analyses could be conducted
to identify specific components that contribute to numerical variability,
and if possible correct them accordingly. Besides, in cases where
instability is inherent to the analysis, sampling results distributions
through numerical perturbations might improve stability, as explored
in~\cite{kiar2021data}. Finally, pipeline-specific statistical corrections
might be envisaged to account for within-tool software variability.
